---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(viridis)
library(scales)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions  <- tidytuesdayR::tt_load('2020-09-22')$expeditions
  
  glimpse(expeditions)
  
 exp1_summary <- expeditions %>%
  mutate(peak = fct_lump(peak_name, n = 15, other_level = "Other")) %>%
  filter(peak != "Other") %>%
  count(peak, name = "n") %>%
  mutate(peak = fct_reorder(peak, n))
 
 expeditions <- tidytuesdayR::tt_load('2020-09-22')$expeditions

# Prepare the data
exp1_summary <- expeditions |>
  mutate(peak = forcats::fct_lump(peak_name, n = 15, other_level = "Other")) |>
  filter(peak != "Other") |>
  count(peak, season, name = "n") |>
  group_by(peak) |>
  mutate(total = sum(n)) |>
  ungroup() |>
  mutate(peak = forcats::fct_reorder(peak, total))

expeditions <- tidytuesdayR::tt_load('2020-09-22')$expeditions

exp1_summary <- expeditions |>
  mutate(peak = forcats::fct_lump(peak_name, n = 15, other_level = "Other")) |>
  filter(peak != "Other") |>
  count(peak, season, name = "n") |>
  group_by(peak) |>
  mutate(total = sum(n)) |>
  ungroup() |>
  mutate(peak = forcats::fct_reorder(peak, total))

expeditions <- tidytuesdayR::tt_load('2020-09-22')$expeditions

# Data preparation
exp1_summary <- expeditions |>
  mutate(peak = forcats::fct_lump(peak_name, n = 15, other_level = "Other")) |>
  filter(peak != "Other") |>
  count(peak, season, name = "n") |>
  group_by(peak) |>
  mutate(total = sum(n)) |>
  ungroup() |>
  mutate(peak = forcats::fct_reorder(peak, total))

expeditions <- tidytuesdayR::tt_load('2020-09-22')$expeditions

exp1_summary <- expeditions |>
  mutate(peak = forcats::fct_lump(peak_name, n = 15, other_level = "Other")) |>
  filter(peak != "Other") |>
  count(peak, season, name = "n") |>
  group_by(peak) |>
  mutate(total = sum(n)) |>
  ungroup() |>
  mutate(peak = forcats::fct_reorder(peak, total))

exp1_summary |>
  ggplot(aes(x = peak, y = n, fill = season)) +
  geom_col() +
  coord_flip() +
  scale_fill_viridis_d() +
  labs(
    title = "The 15 most popular peaks stacked by season of expedition",
    x = NULL,
    y = "Number of expeditions",
    fill = "season"
  ) +
  theme_light() +
  theme(
    legend.position = "bottom",
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title.x = element_text(margin = margin(t = 10))
  )

```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
 # Load dataset
phd_by_field <- tidytuesdayR::tt_load('2019-02-19')$phd_by_field

# Summarize number of PhDs by year and broad field
phd_summary <- phd_by_field |>
  group_by(year, broad_field) |>
  summarise(n = sum(n_phds, na.rm = TRUE), .groups = "drop")

# Plot
phd_summary |>
  ggplot(aes(x = year, y = n, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = scales::pretty_breaks()) +
  scale_y_continuous(labels = scales::comma_format()) +
  scale_color_brewer(palette = "Dark2", name = "Broad field") +
  labs(
    title = "Number of awarded Ph.D.-s in the US by year",
    x = NULL,
    y = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right"
  )

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.


```{r commute-plot, message=FALSE, warning=FALSE}
# Load data
tt_data <- tidytuesdayR::tt_load(2019, week = 45)
commute <- tt_data$commute

# Add region and state abbreviation
commute <- commute |>
  dplyr::mutate(
    region = state.region[match(state, state.name)],
    state_abbr = state.abb[match(state, state.name)]
  )

# Rename region levels to match the legend in your image
commute <- commute |>
  dplyr::mutate(
    region = dplyr::case_when(
      is.na(region)        ~ "NA",
      region == "Midwest"  ~ "North Central",
      TRUE                 ~ region
    )
  )

# Order regions to match legend
commute$region <- factor(commute$region, levels = c("North Central", "Northeast", "South", "West", "NA"))

# Aggregate
commute_summary <- commute |>
  dplyr::filter(mode %in% c("Bike", "Walk")) |>
  dplyr::group_by(state, state_abbr, region, mode) |>
  dplyr::summarise(total = sum(n), .groups = "drop") |>
  tidyr::pivot_wider(names_from = mode, values_from = total) |>
  dplyr::rename(walk = Walk, bike = Bike)

# Plot
ggplot(commute_summary, aes(x = walk, y = bike)) +
  geom_point(aes(color = region), size = 2) +
  geom_text(aes(label = state_abbr), size = 3, vjust = -0.5, check_overlap = TRUE) +
  scale_x_log10(labels = scales::comma_format()) +
  scale_y_log10(labels = scales::comma_format()) +
  scale_color_manual(
    values = c(
      "North Central" = "#e41a1c",  # red
      "Northeast"     = "#4daf4a",  # green
      "South"         = "#00ced1",  # cyan
      "West"          = "#984ea3",  # purple
      "NA"            = "#666666"   # gray
    )
  ) +
    labs(
    title = "Number of people walking vs. biking to work in each USA state",
    x = "Number of ppl walking to work (log N)",
    y = "Number of ppl biking to work (log N)",
    color = "State region"
  )


```
